---
title: "Statisitcal Programming - 2020 - Project 1"
output: rmarkdown::html_document
author: Add your name here
---

Due on Monday, November 2nd by 5:00 pm (17:00), UK local time.

Before you start be sure to read through *all* of the rules and instructions in the `README.md`.

<br/>

### Data

```{r}
library(tidyverse)
library(ggrepel)
library(patchwork)
f1 = readRDS(file="data/f1.rds")
```


<br/>

### Task 1 - Tidy the data

#### Write up

In this task we mainly utilize the `hoist()` function to extract the relevant information
from the dataset before finishing with the `mutate()` function to convert the data
into the appropriate types. Note that the positions “R” (retired), “D” (disqualified), 
“E” (excluded),  “W” (withdrawn), “F” (failed to qualify) and “N” (not classified) 
provided in the original data have all been intentionally coerced into `NA` values
since all of these positions correspond to a driver failing to complete a race.

#### Code

```{r task1}
f1_tbl = tibble(f1_data = f1) %>%
  # ------------------ Pulling out relevant information ------------------ #
  unnest_wider(f1_data) %>%
  hoist(
    MRData,
    races = c('RaceTable', 'Races')
  ) %>% select(races) %>%
  unnest_longer(races) %>%
  hoist(
    races,
    race_name = c('raceName'),
    round = c('round'),
    date = (c('date')),
    results = c('Results')
  ) %>% select(-races) %>%
  unnest_longer(results) %>%
  hoist(
    results,
    first_name = c('Driver', 'givenName'),
    last_name = c('Driver', 'familyName'),
    constructor = c('Constructor', 'name'),
    position = c('positionText'),
    points = c('points')
    ) %>%
  # --------------- Mutating data to be of appropriate type -------------- #
  mutate(
    round = as.integer(round),
    date = as.Date(date),
    driver = paste(first_name, last_name),
    # Suppressing warning NA coercion is desired
    position = suppressWarnings(as.integer(position)),
    points = as.integer(points)
  ) %>% 
  select(race_name, round, date, driver, constructor, position, points)

print(head(f1_tbl, n = 10L))
```

<br/>


### Task 2 - Drivers' Championship

#### Write up

We begin our table creation by first selecting only the relevant columns. We then
proceed to calculate the F1 rankings for the 2019 season before pivoting wider
to make each row represent the positions of a single driver across the season.
Our table columns were renamed as well as the points tally relocated to the end
of the table in order improve the overall readability of the presentation.

#### Code

```{r task2}
# Defining new column names for easy reading
rankings_colnames = c('Driver', 'AUS', 'BHR', 'CHN', 'AZE', 'ESP', 'MON', 'CAN', 'FRA', 
             'AUT', 'GBR', 'GER', 'HUN', 'BEL', 'ITA', 'SIN', 'RUS', 'JPN', 'MEX',
             'USA', 'BRA', 'ABU', 'Points')

rankings_tbl = f1_tbl %>% 
  select(race_name, driver, position, points) %>%
  # Getting ranking for 2019 season for each driver
  group_by(driver) %>%
  mutate(points = sum(points)) %>%
  # 1 row = 1 driver
  pivot_wider(names_from = race_name, values_from = position) %>%
  arrange(desc(points)) %>%
  # Making points the last column
  relocate(points, .after = last_col())
  colnames(rankings_tbl) = rankings_colnames

knitr::kable(rankings_tbl)
```

<br/>

### Task 3 - Cumulative Constructors

#### Write up

We, as usual, begin by selecting only the relevant information. The data is structured
such that we have twice as many rows as required due to each constructor/team
having two representatives. We use this to our advantage and compute the sum of
points grouped by `race_name` and `constructor`. This produces duplicate entries in
our dataset which are easily removed using the `distinct()` function yielding
the points scored by each constructor at each race. Naturally, we can now apply
the `cumsum()` function column wise, pivot wider and sort by descending points 
to generate the required table. Again, renaming the columns was performed to 
improve the readability of the table and match the convention used in Task 2.

#### Code

```{r task3}
# Defining new column names for easy reading
team_colnames = c('Constructor', 'AUS', 'BHR', 'CHN', 'AZE', 'ESP', 'MON', 
                         'CAN', 'FRA',  'AUT', 'GBR', 'GER', 'HUN', 'BEL', 'ITA', 
                         'SIN', 'RUS', 'JPN', 'MEX', 'USA', 'BRA', 'ABU')

team_tbl = f1_tbl %>%
  select(race_name, constructor, points) %>%
  # Getting points scored by each constructor at each race
  group_by(constructor, race_name) %>%
  mutate(points = sum(points)) %>%
  distinct()  %>%
  # Performing cumulative sum of points over the season
  group_by(constructor) %>%
  mutate(points = cumsum(points)) %>%
  pivot_wider(names_from = race_name, values_from = points) %>%
  # Abu Dhabi = final race hence apply sorting here
  arrange(desc(`Abu Dhabi Grand Prix`))
colnames(team_tbl) = team_colnames

knitr::kable(team_tbl)
```


<br/>

### Task 4 - Visualization

#### Write up

For Task 4 I chose to tell the story of the 2019 F1 season through visualizing
the Driver's and Constructor's Championship standings as the season progressed.
The data clearly shows a vary stable and dominant season with Mercedes, Ferrari 
and Red Bull occupying the 1st, 2nd and 3rd places respectively for the entirety
of the 2019 season. A similar but slightly more competitive story is told in 
the Driver's championship with Hamilton and Bottas fighting for 1st up until
around the Hungarian Grand Prix (race 12), before Hamilton begins pulling away to win the 
championship. 

Due to the large gap in terms of points between the top performers
and the rest of the pack, I hence decided to only highlight the top 5 drivers and 
top 3 constructors using each constructor's respective hexadecimal colours. I chose
this common colour scheme to permeate through both figures as I believe it
further illustrates the dominance Mercedes had over the 2019 season and allows the
reader to clearly see how the Constructor's Championship figure was derived 
from the Driver's Championship figure. However, the common colour scheme meant I had
to annotate the top 5 drivers in order to distinguish two drivers from the 
same team which I think looks slightly cramped but worth the cost.
A final design decision that was taken was for both figures to share the same
x-axis. This was done to enable the reader to simultaneously observe the how the
standings for both championships updated by each grand prix.

#### Code

```{r, nametask4, fig.height = 10, fig.width = 10}
# Used to define factor and place races in chronological order
race_levels = c("Australian Grand Prix", "Bahrain Grand Prix", "Chinese Grand Prix",
                "Azerbaijan Grand Prix", "Spanish Grand Prix", "Monaco Grand Prix",
                "Canadian Grand Prix", "French Grand Prix", "Austrian Grand Prix",
                "British Grand Prix", "German Grand Prix", "Hungarian Grand Prix",
                "Belgian Grand Prix", "Italian Grand Prix", "Singapore Grand Prix",
                "Russian Grand Prix", "Japanese Grand Prix", "Mexican Grand Prix",
                "United States Grand Prix", "Brazilian Grand Prix", "Abu Dhabi Grand Prix")

# Hexadecimal colours of constructor each driver belongs to
driver_colours = c("Lewis Hamilton" = "#00D2BE", "Valtteri Bottas" = "#00D2BE",
                   "Max Verstappen" = "#0600EF", "Sebastian Vettel" = "#DC0000",
                   "Charles Leclerc" = "#0600EF")

# Hexadecimal colours of constructors
constructor_colours = c("Mercedes" = "#00D2BE", "Ferrari" = "#DC0000",
                        "Red Bull" = "#0600EF")

# Manipulating driver data to be suitable for ggplot
driver_tbl = f1_tbl %>%
  select(race_name, driver, points, constructor) %>%
  # Getting points scored by each driver at each race
  group_by(driver, race_name) %>%
  mutate(points = sum(points)) %>%
  # Performing cumulative sum of points over the season
  group_by(driver) %>%
  mutate(points = cumsum(points)) %>%
  mutate(race_name = factor(race_name, levels = race_levels))

# Creating filtered data to highlight top 5 drivers in figure
driver_tbl_filtered = driver_tbl %>%
  group_by(driver, constructor) %>%
  filter(max(points) > 200) %>%
  ungroup()

# Plotting
driver_plot = ggplot(driver_tbl, aes(x = race_name, y = points)) +
  # Line plots for each driver by grand prix
  geom_line(
    aes(group = driver), 
    data = driver_tbl,
    color = alpha("grey", 0.7)) +
  # Highlighting top 5 drivers
  geom_line(
    aes(group = driver, color = driver),
    data = driver_tbl_filtered) +
  # Annotating top 5 drivers
  geom_text_repel(
    aes(label = driver),
    nudge_x = .1, nudge_y = -3.5, box.padding = 1.3, point.padding = 1.2,
    arrow = arrow(length = unit(0.02, "npc")), 
    segment.alpha = 0.25,
    size = 6,
    data = driver_tbl_filtered[101:105, ]) + 
  # Adding points over lines since grand prixs are discrete
  geom_point(
    aes(group = driver, color = driver),
    size = 2,
    data = driver_tbl_filtered
  ) + 
  # Aesthetic formatting
  theme_minimal() +
  theme(
    text = element_text(size = 16),
    legend.position = "none", 
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    panel.grid.minor = element_blank()) +
  # Using ticks defined in Task 2 & 3
  scale_x_discrete(labels = team_colnames[-1]) + 
  scale_color_manual(values = driver_colours) + 
  labs(
    title = "F1 Driver's Championship Standings by Grand Prix",
    x = "Grand Prix",
    y = "Points"
  ) 

# Manipulating constructor data to be suitable for ggplot
constructor_tbl = f1_tbl %>%
  select(race_name, constructor, points) %>%
  # Getting points scored by each constructor at each race
  group_by(constructor, race_name) %>%
  mutate(points = sum(points)) %>%
  distinct()  %>%
  # Performing cumulative sum of points over the season
  group_by(constructor) %>%
  mutate(points = cumsum(points)) %>%
  mutate(race_name = factor(race_name, levels = race_levels))

# Creating filtered data to highlight top 3 constructors in figure
constructor_tbl_filtered = constructor_tbl %>%
  group_by(constructor) %>%
  filter(max(points) > 200) %>%
  ungroup()

# Plotting
team_plot = ggplot(constructor_tbl, aes(x = race_name, y = points)) +
    # Line plots for each constructor by grand prix
  geom_line(
    aes(group = constructor), 
    data = constructor_tbl,
    color = alpha("grey", 0.7)) +
  # Highlighting top 3 constructors
  geom_line(
    aes(group = constructor, color = constructor),
    data = constructor_tbl_filtered) +
  # Adding points over lines since grand prixs are discrete
  geom_point(
    aes(color = constructor),
    size = 2,
    data = constructor_tbl_filtered
  ) +
  # Aesthetic formatting
  theme_minimal() +
  theme(
    text = element_text(size = 16),
    legend.position = "bottom",
    panel.grid.minor = element_blank()) +
  # Using ticks defined in Task 2 & 3
  scale_x_discrete(labels = team_colnames[-1]) +
  scale_color_manual(values = constructor_colours) +
  labs(
    title = "F1 Constructor's Championship Standings by Grand Prix",
    x = "Grand Prix",
    y = "Points",
    color = "Constructor"
  )

# Displaying figure
fig = driver_plot / team_plot
fig
```

<br/>

