---
title: "Statisitcal Programming - 2020 - Project 1"
output: rmarkdown::html_document
author: Add your name here
---

Due on Monday, November 2nd by 5:00 pm (17:00), UK local time.

Before you start be sure to read through *all* of the rules and instructions in the `README.md`.

<br/>

### Data

```{r}
library(tidyverse)
f1 = readRDS(file="data/f1.rds")
```


<br/>

### Task 1 - Tidy the data

#### Write up

In this task we mainly utilize the hoist() function to extract the relevant information
from the dataset before finishing with the mutate() function to convert the data
into the appropriate type. Note that the postions “R” (retired), “D” (disqualified), 
“E” (excluded),  “W” (withdrawn), “F” (failed to qualify) and “N” (not classified)  
provided in the original data have all been coerced into NA values since all of 
these positions correspond to a driver failing to complete a race.

#### Code

```{r task1}
f1_tbl = tibble(f1_data = f1) %>%
  # ------------------ Pulling out relevant information ------------------ #
  unnest_wider(f1_data) %>%
  hoist(
    MRData,
    races = c('RaceTable', 'Races')
  ) %>% select(races) %>%
  unnest_longer(races) %>%
  hoist(
    races,
    race_name = c('raceName'),
    round = c('round'),
    date = (c('date')),
    results = c('Results')
  ) %>% select(-races) %>%
  unnest_longer(results) %>%
  hoist(
    results,
    first_name = c('Driver', 'givenName'),
    last_name = c('Driver', 'familyName'),
    constructor = c('Constructor', 'name'),
    position = c('positionText'),
    points = c('points')
    ) %>%
  # --------------- Mutating data to be of appropriate type -------------- #
  mutate(
    round = as.integer(round),
    date = as.Date(date),
    driver = paste(first_name, last_name),
    # Suppressing warning since this produces desired behaviour
    position = suppressWarnings(as.integer(position)),
    points = as.integer(points)
  ) %>% 
  select(race_name, round, date, driver, constructor, position, points)
f1_tbl
```

<br/>


### Task 2 - Drivers' Championship

#### Write up

<!-- Include write up here -->

#### Code

```{r task2}

```

<br/>

### Task 3 - Cumulative Constructors

#### Write up

<!-- Include write up here -->

#### Code

```{r task3}

```

<br/>

### Task 4 - Visualization

#### Write up

<!-- Include write up here -->

#### Code

```{r task4}

```

<br/>

